\chapter{Introduction}
\label{ch:introduction}

\section{Motivation}
\label{sec:motivation}

The industrial sector relies extensively on visually rich documents (VRDs), such as dense engineering datasheets and operating manuals, to communicate critical hardware specifications. These documents are characterized by formidable structural complexities, including embedded engineering graphs, multi-column tables, and hierarchical semantic dependencies. Historically, organizations have depended on manual data entry or simplistic Optical Character Recognition (OCR) to extract this information, leading to severe context loss and a bottleneck for enterprise digitization.

The recent emergence of Vision-Language Models (VLMs) offers a powerful mechanism for automated document intelligence via zero-shot reasoning. However, when deployed against complex industrial documents, both traditional Layout parsing and modern VLMs exhibit severe failure states. Traditional parsing often destroys the reading order. Conversely, VLMs frequently suffer from "structural hallucination" (a phenomenon where the model misinterprets the internal axes of an embedded engineering graph as tabular grid lines, recursively generating false data to satisfy the perceived geometric constraints) or crash due to maximum token limits when attempting to process ultra-dense hardware tables.

To enable true, automated intelligence, industrial PDFs must first be transformed into a hierarchical, mathematically rigid Markdown representation. While the ultimate objective of enterprise digitization is to produce structured, machine-readable JSON payloads, attempting to map raw images directly into rigid JSON schemas frequently triggers catastrophic token exhaustion in generative models. Therefore, this thesis motivates the use of dense, visually representative Markdown as a necessary Intermediate Representation (IR). By first harvesting the document into high-density Markdown, the pipeline radically reduces the token burden, decoupling visual alignment from semantic logic.

Furthermore, industrial environments present unique deployment constraints. Technical datasheets frequently contain proprietary intellectual property, precluding the use of cloud-based APIs (e.g., OpenAI's GPT-4o) due to data privacy regulations. Concurrently, the domain volatility of industrial equipment means that pre-defining rigid database schemas is impossible; a dynamic schema discovery mechanism is required. This thesis, therefore, restricts its hardware footprint to the capacity of a single consumer-grade 24 GB GPU (e.g., NVIDIA RTX 3090), ensuring that the entire pipeline—from dynamic schema synthesis to dual-model reasoning—can execute entirely on secure edge servers.

\section{Aim of this thesis}
\label{sec:aim}

This thesis addresses the limitations of mapping unstructured VRDs to structured data by proposing a novel, multi-modal ingestion framework. The primary objective is to engineer a deterministic pipeline that converts a complex industrial PDF into structural Markdown, eliminating visual hallucination loops.

This objective is systematically achieved through the following technical contributions:

\textbf{1. Ensemble Layout Detection and Verification:}
This research evaluates the application of DocLayout-YOLO transformers to isolate geometric document regions. Rather than relying on generative approximation, the pipeline employs a combination of native PDF primitive extraction and VLM-based reading order verification, providing high-fidelity spatial sequencing. 

\textbf{2. Algorithmic Routing and Two-Pass Extraction:}
To process tabular data with differing complexities, this thesis engineers an intelligent scoring Router. Utilizing key-value gap ratios and vector line density, the system dynamically routes grids to deterministic extractors or specialist VLMs. Crucially, the research introduces a "Two-Pass Markdown Harvest" extraction strategy. To bypass VLM token output bottlenecks and system crashes on dense datasheets, the first pass converts visual tables into high-density raw Markdown. A secondary text-only refinement pass then maps this raw data into Pydantic-validated JSON, guaranteeing full parameter recall.

\textbf{3. Semantic Context and Knowledge Graphs:}
This research introduces a Hierarchical Knowledge Graph (HKG) architecture to preserve the multi-modal "connective tissue" of industrial documents. By implementing a 3-pass construction algorithm with breadcrumb propagation, the pipeline ensures that every extracted text chunk carries its full semantic provenance, optimizing it for downstream RAG applications.

\textbf{4. Parameter-Efficient Edge Deployment:}
To ensure the multi-modal framework executes without relying on latency-heavy proprietary APIs, this thesis utilizes Knowledge Distillation. Using Low-Rank Adaptation (LoRA) and the Unsloth framework, the cognitive load is split into a Dual-Student Architecture. Visual mapping tasks are distilled into an 11-Billion parameter Llama-Vision model, while logical data extraction capabilities are distilled into a 7-Billion parameter Qwen2.5-VL student model. This enables high-throughput execution on a single consumer GPU without triggering CUDA memory failures.

An overview of this proposed architecture is illustrated in Figure \ref{fig:overall_architecture}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=2.5cm and 2cm,
    box/.style={rectangle, draw, thick, align=center, minimum width=3.5cm, minimum height=1.2cm, fill=white, rounded corners=0.1cm},
    arrow/.style={->, thick, >=stealth}
]

\node[box, fill=gray!10] (pdf) {Raw Industrial\\PDF Document};
\node[box, right=2cm of pdf, fill=cyan!10] (layout) {\textbf{DocLayout-YOLO} \\ \& Topological Reading Order \\ \textit{(Section 3.2)}};
\node[box, below=1.5cm of layout, fill=yellow!10] (router) {\textbf{Algorithmic Routing Engine} \\ Heuristic Geometric Scoring \\ \textit{(Section 4.1)}};

\node[box, below left=1.5cm and 0.5cm of router, fill=blue!10] (det) {\textbf{Deterministic Analyzers} \\ Grid Extractor \\ \textit{(Section 4.2)}};
\node[box, below right=1.5cm and 0.5cm of router, fill=green!10] (vlm) {\textbf{Two-Pass Markdown Harvest} \\ Dual-Student Extractors \\ \textit{(Section 4.3 \& Chapter 6)}};

\node[box, below=4cm of router, fill=orange!10] (qa) {\textbf{QA \& Coverage Layer} \\ LangExtract AST Validation \\ \textit{(Section 4.4)}};
\node[box, below=1.5cm of qa, fill=purple!10] (hkg) {\textbf{Hierarchical Knowledge Graph} \\ Serialized IR Generation \\ \textit{(Chapter 5)}};

\draw[arrow] (pdf) -- (layout);
\draw[arrow] (layout) -- (router);
\draw[arrow] (router) -| node[above, align=center, xshift=-1cm] {Ruled\\(Simple Grids)} (det);
\draw[arrow] (router) -| node[above, align=center, xshift=1cm] {Complex/KV\\(Dense Geometry)} (vlm);
\draw[arrow] (det.south) |- (qa.west);
\draw[arrow] (vlm.south) |- (qa.east);
\draw[arrow] (qa) -- (hkg);

\end{tikzpicture}
\caption{The Overall Document Intelligence Pipeline Architecture.}
\label{fig:overall_architecture}
\end{figure}



\section{Structure of this thesis}
\label{sec:structure}

This thesis is organized sequentially, following the flow of geometric data through the multi-modal extraction pipeline:

\begin{itemize}
    \item \textbf{Chapter 2: Background and Related Work} examines the evolution of Vision-Language Models, the geometric limitations of Table Structure Recognition (TSR), and the computational requirements of Parameter-Efficient Fine-Tuning (PEFT).

    \item \textbf{Chapter 3: Layout Intelligence and Reading Order} details the foundational architecture of the ingestion phase, formalizing the DocLayout-YOLO detection and the hybrid use of native PDF primitives and VLM verification.

    \item \textbf{Chapter 4: The Routing Engine and Structured Extraction} analyzes the table classification logic and documents the transition to Pydantic-validated JSON extraction for specialized industrial taxonomies.

    \item \textbf{Chapter 5: Semantic Context and Knowledge Graphs} describes the 3-pass HKG construction algorithm and the Semantic Pre-Masking technique used to resolve table hallucinations.

    \item \textbf{Chapter 6: Custom Model Distillation (PEFT)} documents the precise hyperparameters and empirical training metrics required to distill the Qwen2-VL student model.

    \item \textbf{Chapter 7: Experimental Evaluation} presents the performance metrics of the pipeline across industrial datasets, comparing accuracy, latency, and extraction fidelity against baseline systems.

    \item \textbf{Chapter 8: Conclusions and Future Work} summarizes the research contributions and outlines future directions for scale-up and model specialization.
\end{itemize}