\chapter{Background and Literature Review}
\label{ch:background}

This chapter establishes the theoretical and technical foundation for the pipeline developed in this thesis. It traces the progression of document layout analysis from early heuristic methods to modern transformer-based detectors, reviews the table structure recognition architectures the pipeline builds upon, introduces the language and vision-language models involved in routing and distillation, and covers the parameter-efficient fine-tuning techniques that make local deployment feasible. Finally, it introduces the Retrieval-Augmented Generation paradigm that contextualizes the pipeline's downstream purpose.

\section{The Evolution of Document Layout Analysis}
\label{sec:layout_evolution}

The foundational task in processing Visually Rich Documents (VRDs) is Document Layout Analysis (DLA), which is the geometric identification of semantically distinct regions such as titles, running text, tables, figures, and captions \citep{docai_survey}. Precise region isolation is a prerequisite for all downstream extraction steps; errors at this stage propagate through the entire pipeline without correction.

\subsection{Heuristic and Convolutional Paradigms}

Early DLA relied on bottom-up heuristic procedures, such as horizontal projection profiles and Connected Component Analysis \citep{xycut_ogorman}. While these approaches are computationally lightweight, they exhibit significant failure rates on semi-structured industrial documents that contain multi-column layouts, bordering lines shared between tables and adjacent text, or non-uniform page margins.

The transition to Convolutional Neural Networks brought a substantial improvement. Architectures such as Faster R-CNN framed DLA as an object detection task, using a Region Proposal Network to generate candidate box hypotheses that are refined by convolutional feature pooling \citep{publaynet, yolo}. However, the locally receptive nature of convolutional kernels limits the model's ability to capture spatial relationships that extend across an entire page, such as correctly bounding a schematic that spans two columns.

\subsection{Detection Transformers (DETR)}

The Detection Transformer (DETR) \citep{detr_carion} addressed the global-context limitation of CNNs by applying a standard encoder--decoder transformer directly to the flattened output of a convolutional backbone. The encoder's self-attention mechanism enables each position to attend to all other positions, allowing the model to reason about global spatial distributions. This is particularly valuable in document analysis, where the relationship between a table caption and its parent table may span a large vertical distance.

As depicted in Figure~\ref{fig:detr_evolution}, DETR predicts a fixed set of bounding boxes and class labels in a single forward pass, dispensing with hand-engineered anchor boxes and Non-Maximum Suppression post-processing.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 1cm,
    box/.style={rectangle, draw, thick, align=center, minimum width=2.5cm, minimum height=1cm, rounded corners=0.1cm},
    arrow/.style={->, thick, >=stealth}
]

\node[box, fill=gray!10] (img) {Document Page Tensor\\$1024 \times 1024$};
\node[box, below=1cm of img, fill=blue!10] (cnn) {CNN Backbone\\Feature Extraction};
\node[box, below=1cm of cnn, fill=red!10] (encoder) {Transformer Encoder\\(Global Self-Attention)};
\node[box, below=1cm of encoder, fill=orange!10] (decoder) {Transformer Decoder\\(Object Queries)};
\node[box, below=1cm of decoder, fill=green!10] (ffn) {Feed-Forward Heads\\(BBox \& Class)};

\draw[arrow] (img) -- (cnn);
\draw[arrow] (cnn) -- (encoder);
\draw[arrow] (encoder) -- (decoder);
\draw[arrow] (decoder) -- (ffn);

\end{tikzpicture}
\caption{The Detection Transformer (DETR) architecture \citep{detr_carion}. By replacing the Region Proposal Network with a globally attentive encoder--decoder, DETR predicts bounding boxes without hand-tuned anchors.}
\label{fig:detr_evolution}
\end{figure}

\subsection{DocLayout-YOLO: The Deployed Layout Model}
\label{subsec:doclayout_yolo}

The pipeline described in this thesis uses \textit{DocLayout-YOLO} \citep{doclayout_yolo} as its primary layout detection engine. To contextually position this model, it is essential to understand its lineage. While the Detection Transformer (DETR) shifted the paradigm toward global representation without anchor boxes, subsequent models like RT-DETR \citep{rtdetr_lv} optimized this architecture for real-time inference by refining the multi-scale transformer encoder. Concurrently, the 'You Only Look Once' (YOLO) family evolved towards anchor-free architectures. DocLayout-YOLO builds deeply on this parallel track, specifically deriving from the YOLOv10 backbone. YOLOv10 is a state-of-the-art anchor-free detector that fundamentally eliminates the need for computationally expensive Non-Maximum Suppression (NMS) post-processing by employing consistent dual assignments during training. This results in a mathematically pristine, end-to-end detection system with massive reductions in latency.

DocLayout-YOLO adapts this YOLOv10 architecture specifically for document analysis via a Global-to-Local (G2L) perception module. This allows the model to capture whole-page structural dependencies (e.g., recognizing that a document is split into two macro-columns) while simultaneously isolating highly granular boundaries (e.g., distinguishing a table's footnote from the table body). Pre-trained on a corpus of over 10 million document images (including PubLayNet \citep{publaynet} and DocBank \citep{docbank}), the network exhibits profound prior knowledge of document semantics. This ensures superior localization accuracy compared to RT-DETR or local CNNs, making it the optimal, low-latency backbone for a pipeline running locally.

\subsection{LayoutLMv3 for Token Classification}
\label{subsec:layoutlmv3}

Beyond region-level detection, the pipeline employs \textit{LayoutLMv3} \citep{layoutlmv3} for token-level semantic classification. LayoutLMv3 extends the BERT-style pre-training paradigm to the multimodal document setting by jointly masking text tokens and image patches during pre-training. This unified masking objective produces representations that align visual features (e.g., font rendering, spatial position) with their corresponding text tokens.

In practice, LayoutLMv3 receives a page image together with the token sequence and their bounding box coordinates, and predicts a semantic label for each text token. This token-level output complements the region-level predictions from DocLayout-YOLO, enabling finer-grained distinctions such as identifying running headers that visually resemble body text.

\section{Native PDF Primitives and Text Extraction}
\label{sec:pdfplumber}

A defining characteristic of industrial datasheets is that they are typically generated programmatically from CAD or ERP systems, meaning the underlying PDF file contains a native, vector-level text stream directly embedded in the page description language (e.g., via tools like \texttt{pdfplumber} \citep{pdfplumber}). This is in contrast to scanned documents, which require rasterization and Optical Character Recognition \citep{donut, nougat}.

The pipeline exploits this property by securely parsing the internal PDF content stream to retrieve precise character-level bounding boxes, font names, font sizes, bold and italic flags, and color values without any rasterization step. This native extraction has two practical advantages over advanced OCR approaches (like Tesseract \citep{tesseract} or docTR \citep{doctr}): it is mathematically exact (no transcription errors for machine-generated text), and it provides structural font metadata that is entirely unrecoverable from pixel-level inference.

These font signatures are used downstream by a \textit{FontAnalyzer} component that infers the document's style hierarchy, distinguishing body text from section headings and titles, by computing frequency distributions of font size and weight across the first few pages. The resulting \textit{DocumentStyleSheet} serves as a deterministic fallback and as a grounding signal that can override potentially hallucinated VLM-hypothesized styles.

\section{Table Structure Recognition}
\label{sec:tsr_limits}

Once layout detection has localized table regions, the internal row--column grid must be reconstructed into a machine-readable form. This task, known as Table Structure Recognition (TSR), is considerably more difficult than layout detection because many industrial tables lack explicit, continuous grid lines; alignment is instead implied by whitespace alone.

\subsection{The Heuristic and Deep Learning Eras of TSR}
\label{subsec:tsr_evolution}

The challenge of TSR has historically been approached through two wildly divergent paradigms: bottom-up heuristics and top-down deep learning.

Initially, heuristic systems heavily utilized the X-Y Cut algorithm. X-Y Cut functions by creating horizontal and vertical projection profiles (histograms of ink density) across the isolated table region. Troughs in the density profile theoretically indicate the empty spaces forming rows and columns. While highly deterministic, X-Y Cut is exceptionally brittle. If an industrial table cell contains a multi-line paragraph that overlaps vertically with text in an adjacent column (or features spanning header cells), the projection profile becomes scrambled, causing the algorithm to merge distinct columns or shatter single rows into meaningless fragments.

This brittleness instigated the shift towards deep learning. Early Convolutional Neural Network (CNN) approaches attempted to adapt pixel-masking algorithms (like Mask R-CNN \citep{matterport_maskrcnn}) to identify cell boundaries. However, tables are fundamentally grid topologies, not arbitrary organic shapes. CNNs inherently struggle with the rigid topological graphs required for TSR because their receptive fields often fail to learn that a vertical ruling line on the far left of an image must mathematically align with a cell boundary spanning the entire image height.

To resolve this topological constraint, research heavily pivoted toward Graph Neural Networks (GNNs). In GNN architectures for TSR (such as TGRNet), document tokens (detected via OCR) are treated as nodes in a fully connected spatial graph. The neural network's objective is to prune edges between nodes that do not belong to the same logical row or column. While GNNs provide excellent topological reasoning, their computational complexity scales quadratically with the number of text tokens. A dense industrial parameter table containing 2,000 words forces the GNN to evaluate 4,000,000 potential edge alignments, triggering massive Out-Of-Memory (OOM) failures on standard edge hardware.

\subsection{Encoder-Only vs Causal Decoders for Spatial Reasoning}
\label{subsec:encoder_vs_decoder}

The modern era of TSR is entirely dominated by the Transformer architecture. However, understanding the frequent catastrophic failures of recent Vision-Language Models (VLMs) like GPT-4o requires a strict delineation between Encoder-Only architectures and Causal Decoder architectures.

Encoder-only models, such as LayoutLM and the Table Transformer (TATR), digest the entire image and its token sequence simultaneously. Because the bidirectional self-attention mathematical formulation ($Attention(Q,K,V) = softmax(\frac{QK^T}{\sqrt{d_k}})V$) permits every token to cross-reference every other token synchronously, the model excels at discrete spatial classification tasks. It can globally perceive that a "Voltage" header column spans the entire table length.

Conversely, Causal Decoders (such as Llama-Vision, Qwen-VL, and the GPT-4 family) are fundamentally bound by the autoregressive masking objective. These models generate output tokens sequentially, conditioned exclusively on previously generated tokens ($P(w_t | w_{1:t-1})$). This means that when a VLM is tasked with extracting a 50-row table directly into JSON, it must generate the JSON keys and values linearly. By row 45, the mathematical "memory" of the absolute spatial coordinates of the table header (which was pushed outside the immediate sequence attention horizon due to context limits) degrades severely.

This degradation manifests uniquely as "structural hallucination." Because the decoder cannot bidirectionally "look back" down the image grid while generating text, it begins to guess the column alignment based purely on semantic text probabilities rather than empirical geometric grounding. If an embedded engineering schematic features coordinate tick marks, the autoregressive decoder will hallucinate those ticks as tabular row values because the visual signal forces generation without the geometric restraint of a rigid spatial encoder. This fundamental flaw in Causal Decoders necessitates the dual-pass architectural routing mechanism engineered in Chapters \ref{ch:routing_vlm} and \ref{ch:masking}.

\subsection{The Table Transformer (TATR)}

The Table Transformer (TATR) \citep{tatr}, introduced by Smock et al., applies the DETR detection paradigm to the internal structure of a table crop. It predicts coordinate matrices for rows, columns, and spanning cells directly from the isolated image region, achieving high accuracy on clean, well-separated grids.

However, TATR exhibits a specific failure mode that is particularly prevalent in industrial datasheets: when the table region contains an embedded engineering schematic --- a wiring diagram, a Cartesian timing plot, or a dimensional drawing --- the dense internal line structure of that graphic is interpreted by TATR as table grid lines, causing it to hallucinate rows and columns within the schematic. This phenomenon and the engineering solution developed to address it are described in detail in Chapter~\ref{ch:tsr_masking}.

\section{Vision-Language Models}
\label{sec:vlms}

For table regions where no physical grid lines exist and gap-ratio scoring cannot reliably confirm a key-value structure, the pipeline delegates extraction to a Vision-Language Model (VLM). Modern VLMs encode a page image through a vision encoder and pass the resulting patch embeddings, together with a text prompt, to an autoregressive language decoder \citep{qwen_vl, gpt4v}. The decoder can then generate structured output such as tables or JSON by treating the extraction task as conditional text generation \citep{json_mode}.

The primary risk of VLM-based extraction is \textit{structural hallucination}: the model generates output that is syntactically coherent but semantically inconsistent with the source document \citep{docvqa}, often by misreading numeric labels inside an embedded graphic as table cell values. Furthermore, VLM architectures suffer from strict maximum context generation limits. When forced to map an ultra-dense, multi-page technical catalog directly into structured JSON, the process routinely triggers memory constraints or abruptly truncates output, resulting in completely invalid extraction formats. The Two-Pass Markdown Harvest and routing layers described in subsequent chapters are designed to confine VLM calls to raw unconstrained text generation (Markdown) to sidestep these hardware limitations before structural refinement.

\section{Knowledge Distillation and Parameter-Efficient Fine-Tuning}
\label{sec:peft}

Deploying a capable VLM for layout verification and grid extraction on consumer hardware requires specializing a smaller model to execute the narrow set of tasks the pipeline requires, rather than relying on a large general-purpose API. Knowledge Distillation \citep{knowledge_distill} provides the formal framework for this: a larger, more capable \textit{teacher} model generates training examples (input--output pairs), which are used to train a smaller \textit{student} model to approximate the teacher's behavior on the target task.

\subsection{Hardware Constraints and Edge Deployment}
\label{subsec:hardware_constraints}

A foundational design principle of this thesis is the constraint of execution to a single consumer-grade 24 GB VRAM GPU (such as the NVIDIA RTX 3090). This constraint is not merely an academic exercise; it reflects the strict realities of industrial deployment. Engineering datasheets, electrical schematics, and enterprise operational manuals frequently contain proprietary intellectual property and classified operational parameters. Transmitting this unredacted technical data to third-party cloud LLM APIs (e.g., OpenAI, Anthropic) is strictly prohibited under enterprise data sovereignty definitions, corporate NDA architectures, and privacy regulations.

Consequently, modern document intelligence pipelines must operate entirely intra-network, often deployed on isolated, air-gapped edge servers directly on the manufacturing floor. Bounding the memory requirement to 24 GB guarantees the pipeline avoids the prohibitive capital expenditure of clustered 8xA100 enterprise server racks while maintaining deterministic, low-latency edge isolation limits.

\subsection{Low-Rank Adaptation (LoRA)}

Full fine-tuning of an 11-billion parameter model requires storing gradient tensors whose memory footprint is comparable to or larger than the model weights themselves, placing it far beyond the capacity of a single 24 GB GPU. Low-Rank Adaptation (LoRA) \citep{lora} resolves this by freezing the pre-trained weight matrix $W_0 \in \mathbb{R}^{d \times d}$ entirely and injecting two low-rank factor matrices $A \in \mathbb{R}^{r \times d}$ and $B \in \mathbb{R}^{d \times r}$ alongside it, where $r \ll d$. The modified forward computation adds the low-rank update to the frozen weight output:

\begin{equation}
h = W_0 x + \Delta W x = W_0 x + B A x
\end{equation}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm,
    box/.style={rectangle, draw, thick, align=center, minimum width=2.5cm, minimum height=1cm, rounded corners=0.1cm},
    arrow/.style={->, thick, >=stealth}
]

\node[box, fill=red!10, minimum width=4cm, minimum height=1.5cm] (w0) {Pre-trained Weight Matrix\\$W_0 \in \mathbb{R}^{d \times d}$ (Frozen)};

\node[box, right=2cm of w0, fill=blue!10, minimum width=1.5cm, minimum height=2.5cm] (a) {$A \in \mathbb{R}^{r \times d}$};
\node[box, right=2.5cm of a, fill=green!10, minimum width=2.5cm, minimum height=1.5cm] (b) {$B \in \mathbb{R}^{d \times r}$};

\draw[arrow] (a) -- node[above] {$\times$} (b);

\node[below=0.5cm of a, align=center] {Trainable\\Down-projection ($r \ll d$)};
\node[below=0.5cm of b, align=center] {Trainable\\Up-projection};

\node[above=0.2cm of w0] {$h = W_0 x$};
\node[above left=0.6cm and -1.5cm of b] {$+ \Delta W x$};

\end{tikzpicture}
\caption{LoRA architecture \citep{lora}. Freezing $W_0$ and training the low-rank factors $A$ and $B$ reduces trainable parameters by approximately 98\%, making 11B-parameter model fine-tuning feasible on a single 24 GB GPU.}
\label{fig:lora_architecture}
\end{figure}

By constraining the rank, for instance $r = 16$ as used in this work, the number of trainable parameters drops from $d^2$ to $2rd$, a reduction of roughly 98\% for typical transformer hidden dimensions \citep{qlora}. Training is further accelerated through the \texttt{Unsloth} framework \citep{unsloth}, which replaces PyTorch's standard computational graph with memory-mapped Triton/CUDA kernels optimized specifically for LoRA gradient computation, reducing peak VRAM usage by approximately 60\% compared to a standard PEFT \citep{peft_houlsby} implementation.

\section{Retrieval-Augmented Generation}
\label{sec:rag}

The structured Markdown and graph output produced by this pipeline are intended for downstream consumption by a Retrieval-Augmented Generation (RAG) system \citep{rag_lewis}. In the RAG paradigm, a language model answers queries by first retrieving relevant passages from an indexed document corpus, then conditioning its generation on those passages. The quality of the retrieval step --- and therefore the quality of the generated answers --- depends critically on how accurately the document corpus has been chunked and indexed.

A flat OCR transcript of an industrial datasheet, with destroyed reading order and collapsed tables, produces low-quality retrieval chunks that cause the downstream model to hallucinate specifications or misattribute parameters to the wrong product. By contrast, the Hierarchical Knowledge Graph (HKG) produced by this pipeline preserves section context, maintains table integrity, and resolves cross-references, yielding retrieval chunks that carry unambiguous semantic provenance. This is the principal motivation for the graph-native representation that will be comprehensively detailed in Chapter 5.

