\chapter{Layout Intelligence and Reading Order}
\label{ch:layout_reading}

The fundamental bottleneck in transforming industrial Visually Rich Documents (VRDs) into hierarchical Markdown is the catastrophic loss of spatial geometry during raw text extraction. Commercial OCR engines operate primarily via horizontal sweeping, inherently destroying the semantic flow of multi-column layouts, tables, and nested footnotes. 

This chapter details the foundational architecture of the High-Fidelity Ingestion Pipeline developed within this research. The methodology is executed in three critical geometric phases: bounding box prediction via DocLayout-YOLO \cite{doclayout_yolo} (a real-time detection transformer derivative), native primitive extraction via direct PDF stream parsing \cite{pdfplumber}, and topological reading-order verification utilizing deterministic Recursive XY-Cut whitespace projections \cite{xycut_ogorman} and VLM-based continuity scoring.

\section{System Architecture and Ensemble Flow}
\label{sec:system_architecture}

To ensure deterministic stability across massive engineering datasheets without relying on a monolithic Language Model, the framework utilizes an ensemble pipeline. Specialized visual tasks are delegated to highly local, memory-efficient models.

Figure \ref{fig:architecture_flowchart} illustrates the massive, end-to-end data flow of the ingestion phase, leading directly into the Routing and Masking layers discussed in subsequent chapters.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2cm,
    box/.style={rectangle, draw, thick, align=center, minimum width=2.5cm, minimum height=1.0cm, rounded corners=0.1cm},
    arrow/.style={->, thick, >=stealth}
]

% Nodes
\node[box, fill=gray!10] (pdf) {Raw Industrial PDF};
\node[box, below=1cm of pdf, fill=blue!10] (layout) {Layout Detection\\(DocLayout-YOLO)};
\node[box, left=2cm of layout, fill=red!10] (images) {Raster Image\\Patches};
\node[box, right=2cm of layout, fill=orange!10] (text) {Native Text\\Primitives (pdfplumber)};

\node[box, below=1cm of layout, fill=purple!10] (xycut) {XY-Cut Reading\\Order Algorithm};
\node[box, right=2cm of xycut, fill=yellow!10] (font) {FontAnalyzer \&\\Stylesheet};

\node[box, below=1.5cm of xycut, fill=green!10] (verification) {VLM Continuity\\Verification};

% Connections
\draw[arrow] (pdf) -- (layout);
\draw[arrow] (pdf) -| (images);
\draw[arrow] (pdf) -| (text);

\draw[arrow] (layout) -- (xycut);
\draw[arrow] (text) -- (font);
\draw[arrow] (font) -- (xycut);
\draw[arrow] (xycut) -- (verification);
\draw[arrow] (images) |- (verification);

\end{tikzpicture}
\caption{The Ensemble Architecture of the Layout Detection Phase. The pipeline combines DocLayout-YOLO visual predictions with native font primitives to establish reading order, followed by VLM-based verification.}
\label{fig:architecture_flowchart}
\end{figure}

The document is processed sequentially, page-by-page. For each discrete page, it is simultaneously parsed for precise mathematical primitives (via rendering libraries) and rasterized into a high-resolution padded $1024 \times 1024$ pixel tensor for semantic component classification. This isolated page-level recursive loop is essential to bound active VRAM utilization within hardware limits.

\section{Layout Detection via DocLayout-YOLO}
\label{sec:doclayout_yolo_logic}

Extracting raw text from a dual-column industrial datasheet without spatial context results in a disordered jumble of parameters. To impose architectural boundaries, the pipeline utilizes \textit{DocLayout-YOLO} \cite{doclayout_yolo}, a vision-first transformer derivative specialized for document layout analysis.

As described in Chapter 1, DocLayout-YOLO leverages the YOLOv10 anchor-free paradigm. The initial step is data normalization. The raw PDF page is rasterized at 300 DPI into a standard RGB image. To preserve the aspect ratio and prevent geometric distortion of embedded tables, the image is padded with uniform white space (letterboxing) to reach a perfect square configuration. This square image is then normalized into a standard $1024 \times 1024 \times 3$ input tensor.

The model receives this tensor and directly predicts a fixed set of bounding box proposals $B = \{b_1, b_2, ..., b_n\}$ through its multi-scale feature encoder. Each proposal is parameterized as absolute coordinates relative to the input tensor: $(x_{min}, y_{min}, x_{max}, y_{max})$. An MLP classification head assigns a categorical probability distribution $C_i \in \{ \text{Title, Heading, Text, Table, Figure, Caption, Footer} \}$ to each proposed region.

Figure \ref{fig:doclayout_data_flow} maps the explicit dimensional transformation of the input data as it passes through the DocLayout-YOLO pre-processing and inference modules.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm,
    box/.style={rectangle, draw, thick, align=center, minimum width=3cm, minimum height=1cm, rounded corners=0.1cm},
    arrow/.style={->, thick, >=stealth}
]

\node[box, fill=gray!10] (pdf) {Raw PDF Page\\(Vector)};
\node[box, right=1.2cm of pdf, fill=red!10] (raster) {Rasterization\\300 DPI RGB};
\node[box, right=1.2cm of raster, fill=orange!10] (pad) {Letterbox Padding\\(Preserve Ratio)};
\node[box, below=1cm of pad, fill=blue!10] (tensor) {Normalized Tensor\\$1024 \times 1024 \times 3$};
\node[box, left=1.2cm of tensor, fill=cyan!10] (yolo) {DocLayout-YOLO\\Inference Phase};
\node[box, left=1.2cm of yolo, fill=green!10] (output) {Proposal Set $B$\\$(x, y, C_i)$};

\draw[arrow] (pdf) -- (raster);
\draw[arrow] (raster) -- (pad);
\draw[arrow] (pad) -- (tensor);
\draw[arrow] (tensor) -- (yolo);
\draw[arrow] (yolo) -- (output);

\end{tikzpicture}
\caption{Data Flow Diagram detailing the spatial and tensor transformations required to process an unstructured page into normalized bounding box arrays.}
\label{fig:doclayout_data_flow}
\end{figure}

Due to its advanced dual-assignment training, DocLayout-YOLO minimizes the need for NMS (Non-Maximum Suppression). However, to resolve edge-case overlapping predictions — such as a generic "Figure" box duplicating a specific "Table" box — the pipeline applies a localized Intersection over Union (IoU) filter:
\begin{equation}
\text{IoU}(b_a, b_b) = \frac{\text{Area}(b_a \cap b_b)}{\text{Area}(b_a \cup b_b)}
\end{equation}

\section{Native Primitive Extraction and Font-Based Promotion}
\label{sec:font_analyzer}

While visual layout models are effective for geometric boundaries, they often struggle with fine-grained semantic distinctions, such as separating a bold paragraph from a section heading. The pipeline addresses this by parallelizing detection with native PDF primitive extraction using \textit{pdfplumber}.

\subsection{FontAnalyzer and Stylesheet Hypotheses}
While `pdfplumber` acts as the low-level data extraction mechanism providing raw character matrices, the pipeline employs a distinct, custom \textit{FontAnalyzer} algorithm to make structural sense of that data. For each detected text region localized by the visual backbone, the FontAnalyzer mathematically iterates over the nested `pdfplumber` character stream to extract the dominant "font signature" --- comprising the exact TrueType font name, absolute string size, boldness encoding, and RGB color values \cite{pdfplumber}. By analyzing the frequency distribution of these signatures across the preamble of a document, the pipeline auto-generates a \textit{DocumentStyleSheet} using Pydantic JSON schemas \cite{pydantic}. 

This stylesheet allows for deterministic \textbf{Region Promotion}. For instance, a region initially classified as generic "Text" by the visual model is automatically promoted to a structural "Heading" or "Title" if its font size exceeds 12pt and it possesses a bold weight flag, or if it consists entirely of uppercase characters within a highly specific width tolerance. This hybrid approach leverages the geometric strengths of modern vision transformers \cite{vit} while maintaining the pixel-perfect semantic precision afforded by native document metadata.

\section{VLM reading-order Verification}
\label{sec:vlm_verification}

Recursive XY-Cuts (detailed in Section \ref{sec:xy_cuts}) provide a robust first pass for reading order, but complex layouts with interleaved columns or floating sidebars can still lead to semantic breaks. To ensure the integrity of the downstream extraction, the pipeline incorporates a \textit{VLM Verification} layer.

A specialist Vision-Language Model \cite{qwen25} receives the page image and the proposed reading order manifest. It acts as a "referee," scoring the flow continuity on a scale of 1 to 10 based on semantic and visual coherence \cite{chain_of_thought}. If the score falls below a defined threshold (e.g., indicating interleaved columns that disrupt a sentence), the pipeline flags the page for adaptive re-sorting or triggers a "column-first" processing mode. 

\section{Recursive XY-Cut Reading Order Algorithm}
\label{sec:xy_cuts}

Once semantic bounding boxes are isolated $(B)$ and semantic types are verified, their sequential reading order must be resolved. While western reading traditionally flows top-to-bottom, multi-column industrial layouts invalidate naive coordinate sorting.

To provide a deterministic logical flow, the framework utilizes \textbf{Recursive XY-Cuts} \cite{xycut_ogorman}. The algorithm operates by projecting the bounding boxes onto the X and Y axes, searching for spatial "valleys" of whitespace that exceed a minimum gap threshold $\tau$. By recursively partitioning the page along these valleys, the document is decomposed into a hierarchical tree of regions that naturally preserves the intended human reading order. 
