\chapter{The Routing Engine and VLM Specialists}
\label{ch:routing_vlm}

Once a multi-page industrial document has been successfully processed through the Layout Detection phase, the resulting geometric bounding boxes must be interpreted. For generic text paragraphs, standard text primitives are structurally sufficient. However, tabular data within engineering datasheets exists across a vast spectrum of typological complexity. 

This chapter details the architecture of the Algorithmic Routing Engine developed in this thesis. By computing vector line densities and geometrical word-gap ratios, the router dynamically classifies tabular instances and dispatches them to distinct inferential systems: deterministic grid extractors, key-value specialists, or generative Vision-Language Models (VLMs). Furthermore, this section describes the multi-layered verification strategy consisting of constrained JSON generation and an automated Quality Assurance (QA) loop that triggers fallback extraction strategies upon failure.

\section{The Algorithmic Router}
\label{sec:router_logic}

Industrial tables generally manifest in three structural configurations: Ruled grids (explicitly boxed matrices), Key-Value pairs (implicit 2-column alignments), and Complex grids (multi-lane tables often embedding external graphical structures). Applying a massive generative Vision-Language Model to a perfectly ruled 50-row table incurs unacceptable latency and massive hallucination risk. To understand this risk, it is necessary to distinguish between encoder-only models (like LayoutLM \citep{layoutlmv3}, which excel at discrete token classification) and decoder-only autoregressive models (like Qwen2.5-VL or Llama-Vision). Decoder models generate text sequentially; applying a decoder-only model to process a massive, perfectly structured 50-row visual table forces it to autoregressively "guess" the alignment across thousands of tokens, which directly instigates cross-row hallucination. Conversely, applying an implicit spatial algorithm to a 3-lane technical table fails catastrophically due to the lack of explicit boundaries.

To resolve this, an internal sorting orchestrator, the \textit{TableTypeRouter}, computes mathematical heuristics based on the native topological primitives (words and drawing vectors) residing within a defined geometric bounding box. The orchestrator generates two primary continuous values: $S_{\text{ruled}}$ and $S_{\text{kv}}$.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1cm and 2cm,
    box/.style={rectangle, draw, thick, align=center, minimum width=3cm, minimum height=1cm, rounded corners=0.1cm},
    decision/.style={diamond, draw, thick, align=center, aspect=2.5, minimum width=4cm, inner sep=1pt},
    arrow/.style={->, thick, >=stealth}
]

% Nodes
\node[box] (input) {Table Region $B_{table}$};
\node[box, below=0.8cm of input] (scores) {Compute $S_{ruled}$, $S_{kv}$};
\node[decision, below=0.8cm of scores] (is_ruled) {$S_{ruled} > 0.4$?};
\node[decision, below=1.2cm of is_ruled] (is_kv) {$S_{kv} > 0.8$?};

\node[box, right=2cm of is_ruled, fill=blue!10] (ruled) {Deterministic\\Grid Extractor\\(Sec. 3.1.1)};
\node[box, right=2cm of is_kv, fill=green!10] (vlm) {Two-Pass\\Markdown Harvest\\(Sec. 3.2)};

\node[box, below=1.2cm of is_kv, fill=red!10] (tsr) {Complex TSR\\\& Masking\\(Chapter 5)};

\node[box, below=1cm of vlm, fill=orange!10] (qa) {QA Layer\\(Coverage/Duplication)\\(Sec. 3.3)};
\node[box, below=1cm of qa, fill=cyan!10] (extractor) {Extractor Agent\\(Sec. 3.4)};

% Connections
\draw[arrow] (input) -- (scores);
\draw[arrow] (scores) -- (is_ruled);

\draw[arrow] (is_ruled) -- node[above, font=\small] {Yes} (ruled);
\draw[arrow] (is_ruled) -- node[left, font=\small] {No} (is_kv);

\draw[arrow] (is_kv) -- node[above, font=\small] {Yes} (vlm);
\draw[arrow] (is_kv) -- node[left, font=\small] {No} (tsr);

\draw[arrow] (ruled.east) -- ++(0.8,0) |- (qa.east);
\draw[arrow] (vlm) -- (qa);
\draw[arrow] (qa) -- (extractor);

\draw[arrow] (qa.south) |- ++(0,-0.5) -| ([xshift=-2.5cm]scores.west) |- node[near start, left, font=\small, fill=white, inner sep=2pt] {Fail $\rightarrow$ Retry} (scores.west);

\end{tikzpicture}
\caption{The hierarchical routing and verification loop. Tables are routed based on geometric scores; the extraction is subsequently validated by a QA layer that can trigger fallback attempts.}
\label{fig:router_flow}
\end{figure}

\subsection{Vector Line Density Scoring}
To determine if a tabular bounding box contains a discrete physical grid, the router calculates the cumulative Euclidean length of all valid horizontal and vertical drawing primitives ($D$). Let $B$ represent the table bounding box with width $w$ and height $h$. The perimeter is defined as $P = 2(w + h)$. The ruled score $S_{\text{ruled}}$ is normalized against the bounded perimeter:

\begin{equation}
S_{\text{ruled}} = \min \left( 1.0, \frac{\sum_{d \in D} \text{length}(d)}{2(w + h)} \right)
\end{equation}

If $S_{\text{ruled}} > 0.4$, the bounding box is immediately classified as a physically ruled table, triggering deterministic row-column projection parsing. The threshold value of 0.4 was determined empirically. During validation on a subset of 500 industrial PDFs, it was observed that commercial OCR and rasterization artifacts frequently fragment thin vector lines. A threshold of 0.4 provides the optimal balance, confirming sufficient grid density to indicate a table while maintaining robust algorithmic tolerance against pixel-level line breaks.

\subsection{Maximum Gap Ratio Scoring}
For technical datasheets heavily utilizing implicit Key-Value mappings (e.g., "Operating Voltage | 24 V DC"), physical lines are entirely absent. To detect this 2-column configuration without relying on an LLM, the router analyzes the geometric distribution of the textual nodes.

Let $X$ represent the sorted array of the geometric $X$-centers corresponding to every word bounding box within the table. The continuous gaps $G$ between adjacent words are computed. The maximum topological gap $g_{max}$ and its positional ratio relative to the table width $w$ form the basis of the key-value score $S_{\text{kv}}$:

\begin{equation}
S_{\text{kv}} = 0.7 \left( \frac{g_{max}}{w} \right) + 0.3 \left( B_{factor} \right)
\end{equation}

The balance factor, $B_{factor}$, calculates the symmetric distribution of text nodes residing left or right of $g_{max}$. If the combined score $S_{kv} > 0.8$, it is classified as a strict key-value list. This 0.8 threshold ensures a high-confidence boundary, permitting only perfectly aligned dual-column arrays to bypass complex structural analysis. If a table yields a highly imbalanced distribution or exhibits multiple significant gaps ($n \geq 2$), the router detects a complex multi-metric grid, bypassing the VLM and forcing the table to the Semantic Pre-Masking TSR framework detailed in Chapter 5.

\section{Two-Pass Markdown Harvest and Refinement}
\label{sec:vlm_specialist}

When the router encounters a structure that requires generative reasoning ($S_{\text{kv}} > K_{threshold}$), the region is passed to a Vision-Language Extractor \citep{gpt4v, qwen25}. Unlike general-purpose chat models, the extractor used in this pipeline employs a strict Two-Pass strategy to circumvent hardware constraints and token output bottlenecks.

\subsection{Pass 1: High-Density Markdown Generation}
To mathematically force deterministic grid generation without JSON hallucination loops or truncation, the Local VLM Specialist (Qwen2-VL) is prompted to digest the visual grid strictly into continuous Markdown tables. Because Markdown avoids the deep nesting overhead of JSON (braces, brackets, structural keys), it exhibits approximately 300\% greater token efficiency. This ensures that ultra-dense parameter catalogs can be extracted in a single generation window without triggering CUDA out-of-memory errors on consumer GPUs.

\subsection{Generative Prompting and Temperature Bounds}
\label{subsec:prompting_bounds}

The robustness of the local VLM (Qwen2.5-VL) relies fundamentally on the deterministic architecture of the \texttt{System} Prompt instructions. When a tabular crop is injected into the VLM, a strict sequence of \textbf{Negative Constraints} is prepended to the system context. These directives explicitly command the model: "DO NOT invent column headers if none exist. DO NOT output conversational preamble. DO NOT interpolate table values."

This behavior is enforced via strict \textbf{Temperature Bounds}. In autoregressive language modeling, the temperature variable ($T$) scales the logits before the softmax normalization ($softmax(z_i / T)$), directly altering the probability mass of subsequent tokens. A temperature of $T=1.0$ promotes standard language diversity. However, for tabular grid extraction, we mandate a temperature interval of $T \in [0.1, 0.2]$. This hyper-constrained bound brutally restricts the VLM generation to the absolute most probable structural tokens (e.g., standard Markdown pipes `|` and hyphens `-`), eliminating the creative entropy that typically instigates conversational "hallucinations" inside technical grids.

Furthermore, dynamic \textbf{Few-Shot Demonstration} blocks are injected dynamically depending on the TableTypeRouter's geometric classification. If the table is classified as Key-Value ($S_{kv} > 0.8$), a perfectly formatted target representation is provided dynamically within the prompt context window. The combination of an extremely low temperature scalar and deterministic prompt constraints functionally transforms the Generative VLM into a rigid, non-creative OCR array decoder.

\subsection{Pass 2: Text-to-Schema Refinement}
Once the raw Markdown buffer is harvested, a secondary text-only refinement pass maps the tabular data definitively to the final Pydantic JSON schemas \citep{pydantic}. This pass acts as a structural validator, systematically isolating identity metadata (e.g., Article Number, Manufacturer) from the deep technical specifications, yielding a perfect programmatic mapping that is immune to multi-modal geometric hallucinations.

\section{Extraction Quality Assurance and Adaptive Retry}
\label{sec:qa_layer}

A significant contribution of this architecture is the secondary verification layer that follows every extraction attempt. Rather than assuming the first extraction is correct, the \textit{QA Coordinator} evaluates two precision metrics: \textbf{Coverage} and \textbf{Duplication}.

\subsection{Coverage and Duplication Metrics}
The coverage metric measures the ratio of source text tokens successfully mapped into the resulting structure, drawing heavily on the precise source grounding principles of the LangExtract library \citep{langextract}. If a table contains forty numeric parameters but the VLM only extracts ten, the low coverage score triggers an automated failure because the spatial coordinates of the tokens do not trace back to the structured output.

Conversely, the duplication metric detects instances where the model "stutters" or recursively generates the same row multiple times --- a common failure mode in sequence-to-sequence document models. Let $E$ be the set of extracted records. The duplication score $S_{dup}$ is defined as the frequency of redundant entries normalized by the total count. If $S_{dup} > 0.05$ or Coverage $< 0.90$, the extraction is rejected.

\subsection{Adaptive Fallback Logic}
In the event of a QA failure, the pipeline does not stop. Instead, the coordinator triggers an \textit{Adaptive Fallback}. If a generative VLM attempt failed due to duplication error, the system automatically elevates the inference "temperature" variable (introducing higher probabilistic entropy to break repetitive generation loops). Additionally, it modifies the system prompt to enforce a "chain-of-thought" structure, demanding the model explain the grid sequence step-by-step prior to extraction. In extreme truncation cases, the table is geographically subset into horizontal chunks using DocLayout bounding boxes, drastically expanding the available context horizon. This iterative retry loop substantially increases the extraction success rate on highly complex, semi-structured industrial materials.

\subsection{Eradicating Generative Topological Hallucination}
The most pernicious failure state in visual grid reconstruction occurs when a standard VLM processes a technical table encapsulating a visual diagram (e.g., an embedded sensor wiring schematic). Standard instruction-tuned models recursively attempt to OCR the diagram's internal axes or numeric labels, creating endlessly hallucinated, unsynchronized tabular rows. 

To surgically mitigate this hallucination loop, the "Markdown Harvest" strategy enforces prompt-level constraint directives and aggressive repetition penalties during the decoding phase. Additionally, line-level string deduplication ensures that geometric sameness traps (where the visual encoder endlessly loops on a stationary design element) are neutralized before the secondary schema refinement pass. Empirical evaluation reveals that decoupling raw text extraction from structural data formatting drastically suppresses the latent drive of the model to infinitely parse arbitrary graphical lines.

\section{Dynamic Schema Design and Specialist Heuristics}
\label{sec:schema_design}

The ultimate objective of the Document Intelligence Pipeline is not simply to produce flattened Markdown, but to populate deterministic, highly-structured database records. This requirement is fulfilled by the final autonomous component in the pipeline: the \textit{ExtractorAgent}.

\subsection{Pydantic Schema Instantiation via Discovery Agent}
Once the two-pass routing engine resolves complex tables and the Hierarchical Knowledge Graph provides the semantic grouping (see Chapter \ref{ch:graph_masking}), the resulting multi-modal batches are mapped against strict, strongly-typed object schemas. This thesis implements these mappings using the Pydantic \citep{pydantic} framework. 

Crucially, the exact combination of expected data blocks (e.g. \textit{invoice\_header}, \textit{parameters}, \textit{connectors}) is not hard-coded. Instead, a preliminary \textit{DiscoveryAgent} scans a textual preview of the document. By determining the domain (e.g., Industrial vs. Corporate) and synthesizing contextual hints, the pipeline dynamically constructs a unified model at runtime utilizing Python's \texttt{create\_model} constructor. Pydantic is a powerful data validation framework that forces Python dictionaries to strictly adhere to defined typings (e.g., ensuring a "Voltage" field only accepts floating-point numbers). The DiscoveryAgent effectively acts as a programmatic compiler, analyzing the unstructured headers and passing an array of expected string and float \texttt{kwargs} directly into the Pydantic \texttt{create\_model} function, synthesizing a rigid target object schema dynamically out of thin air.

\subsection{LangExtract AST Analysis and Runtime Construction}
\label{subsec:langextract_ast}

The process of binding the unified string arrays to the dynamic Pydantic blueprint relies fundamentally on the internal architecture of the \texttt{LangExtract} \citep{langextract} library. Unlike archaic string-matching parsers, \texttt{LangExtract} operates on a programmatic level. When the Discovery Agent concludes the schema configuration, the pipeline passes the definitions through an Abstract Syntax Tree (AST) analyzer. This parser reads the dictionary schema structures and procedurally transpiles them into a native Python class definition stored internally in system memory.

The AST runtime generation has severe operational benefits. By transforming dynamic JSON schemas into inherently typed logical expressions (e.g. mapping an arbitrary field \texttt{Max\_Current} strictly to a built-in \texttt{float} primitive representing Amperage), \texttt{LangExtract} inherently attaches recursive tracebacks. When the secondary extraction VLM is prompted with this generated AST blueprint, any syntax deviations triggering a validation failure immediately isolate the precise geometric row and token fault. This ensures deterministic rollback capabilities. Without runtime AST construction, the engine would be forced to execute brittle regular expressions to validate JSON outputs, drastically slowing the multi-stage extraction pipeline and crippling verification precision.

For the primary industrial use case, the pipeline predominantly employs a synthesized \texttt{LibrarianUniversalHardware} schema. Rather than allowing an LLM to arbitrarily invent JSON keys, this structural template forcefully guides the final generation matrix:
\begin{itemize}
    \item \textbf{Identity Nodes}: Enforces strict extraction of Article Numbers (\textit{Art.-Nr.}), Manufacturer names, and Product Titles.
    \item \textbf{Technical Parameters}: Reassembles the harvested Markdown tables into discrete electrical (e.g., operating voltage), mechanical (e.g., dimensions), and environmental (e.g., IP rating) parameter classes.
    \item \textbf{Connector Configurations}: Maps wiring schemas strictly into \textit{Pinout} objects requiring a specific signal, pin number, and function.
\end{itemize}

\subsection{Librarian-Grade Agentic Heuristics}
To guarantee maximum coverage against the Pydantic schemas without relying on blind LLM inference, the \textit{ExtractorAgent} dynamically injects context-specific heuristics based on the document type classification.

For instance, if a document batch is classified as an industrial catalog, the Agent retrieves the \textit{Industrial\_datasheet} instruction payload. This payload acts as an aggressive constraint matrix: "Focus entirely on technical specifications. Extract EVERY row of every parameter table... For EVERY connector (M12, M8, RJ45) list ALL pins with signal and function." Other payloads exist in the system architecture, such as a \textit{Loose\_corporate} payload (focusing merely on entities and abstract summary text) or a \textit{Strict\_financial} payload (enforcing strict ledger balancing). Currently, the pipeline hosts five discrete heuristic payloads, dynamically selected based on the DiscoveryAgent's domain classification.

By combining the rigid schema boundaries of Pydantic with these specialized heuristic prompts, the end-to-end pipeline flawlessly transforms a visual PDF into a granular, production-ready JSON manifest, immune to both spatial hallucination and unconstrained key-generation drift.
