\chapter{Semantic Context and Hierarchical Knowledge Graphs}
\label{ch:graph_masking}

Translating an industrial Visually Rich Document (VRD) into a flat text stream destroys the multi-modal relationships that define its meaning --- the link between a table and its caption, or a specification and its parent section header. To preserve this semantic integrity, the pipeline implements two advanced structural layers: a \textit{Semantic Pre-Masking} algorithm to resolve table hallucinations, and a \textit{Hierarchical Knowledge Graph} (HKG) that encodes the document's topological and semantic provenance.

This chapter details these layers, describing how the pipeline moves beyond layout detection to construct a graph-native representation suitable for high-precision Retrieval-Augmented Generation (RAG).

\section{The Core Visual Hallucination Paradigm}
\label{sec:hallucination_paradigm}

The Table Transformer (TATR) utilizes a Detection Transformer (DETR) architecture rigorously pre-trained on grid structures to decode exact row and column bounding vectors directly from image convolutions \cite{tatr}. The model does not explicitly read text; instead, it predicts structural intersections by observing physical grid lines and implicit alignment whitespace.

However, the purely visual mechanism driving TATR's success triggers catastrophic failure when processing multi-modal industrial datasheets. Consider an electrical specification table wherein the third column dynamically embeds a raster image of a wiring schematic. The schematic inherently contains thousands of high-contrast, linear graphic vectors representing wires, circuit terminals, and Cartesian testing boundaries.

WhenTATR executes inference over the parent table bounding box, its convolutional kernels identically interpret the wiring vectors as legitimate structural table boundaries. This visual interference causes TATR to aggressively over-segment the table, hallucinating hundreds of chaotic, non-existent rows within the wiring schematic to mathematically conform to the diagram's geometry. This behavior totally destroys the legitimate cross-row alignment of the true tabular data.

\section{The Semantic Pre-Masking Algorithm}
\label{sec:pre_masking_logic}

To permanently circumvent generative visual hallucination, this thesis engineers an absolute \textit{Semantic Pre-Masking} algorithm—"absolute" indicating that the intervention performs destructive, irreversible optical redaction on the tensor prior to inference. The fundamental logic dictates that if a visual model hallucinates false rows when observing complex schematics, the mathematical tensor underlying those schematics must be erased from the observation prior to inference execution.

Instead of relying on fragile Large Language Model (LLM) instruction tuning to "ignore the graph", the intervention is purely geometrical and localized directly on the raw image array.

The mechanism relies on the metadata synchronization established during the initial ingestion phase. Specifically, the framework utilizes the `ImagePrimitives` array derived from the native PDF structural tree.

Let $T_{crop}$ denote the high-resolution NumPy pixel tensor array representing the structural crop of the localized Table bounding box, bounded by $B_{table} = (X_{min}, Y_{min}, X_{max}, Y_{max})$.
Let $F$ represent the set of all raster images and `Figure` bounding boxes independently identified via DocLayout-YOLO residing within the parent document.

The algorithm mathematically projects the spatial constraints of any external figures $f_i$ directly onto the Table's localized sub-tensor. Hallucination zones are programmatically recognized by computing the geometric intersection coordinates between any DocLayout-YOLO proposal classified as a \texttt{Figure} and any proposal classified as a \texttt{Table} on the same page. If the absolute bounding coordinates of the figure $B_{figure}$ intersect the table crop $B_{table}$, its relative localized coordinates $(x'_{min}, y'_{min}, x'_{max}, y'_{max})$ are derived dynamically using offset subtraction mapping against the Table's coordinate origin:

\begin{equation}
\begin{split}
x'_{min} &= \max(0, B_{figure}.x_{min} - B_{table}.x_{min}) \\
y'_{min} &= \max(0, B_{figure}.y_{min} - B_{table}.y_{min}) \\
x'_{max} &= \min(B_{table}.x_{max} - B_{table}.x_{min}, B_{figure}.x_{max} - B_{table}.x_{min}) \\
y'_{max} &= \min(B_{table}.y_{max} - B_{table}.y_{min}, B_{figure}.y_{max} - B_{table}.y_{min})
\end{split}
\end{equation}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=2.5cm,
    box/.style={rectangle, draw, thick},
]

% Background grid representing tensor
\draw[step=0.5cm,gray,very thin] (0,0) grid (6,4);
\node[above] at (3,4) {Original Table Tensor $\mathbf{T}_{crop}$};

% Sub-region for graphic
\fill[red!30, opacity=0.8] (3.5, 0.5) rectangle (5.5, 3.5);
\draw[red, thick, dashed] (3.5, 0.5) rectangle (5.5, 3.5);
\node[red, align=center] at (4.5, 2) {Hallucination Zone\\(Embedded Graph)};

% Arrows to masking
\draw[->, ultra thick] (6.5, 2) -- (8.5, 2) node[midway, above] {\textit{Pre-Masking}};

% Result grid
\draw[step=0.5cm,gray,very thin] (9,0) grid (15,4);
\fill[white] (12.5, 0.5) rectangle (14.5, 3.5);
\draw[blue, thick] (12.5, 0.5) rectangle (14.5, 3.5);
\node[blue, align=center] at (13.5, 2) {Obscured Zone\\(255, 255, 255)};
\node[above] at (12,4) {Masked Tensor $\mathbf{T}_{crop}'$};

\end{tikzpicture}
\caption{Geometric visualization of the Semantic Pre-Masking algorithm. Destructive optical obscuration totally eliminates highly-dense engineering schematic vectors prior to Table Transformer (TATR) object detection inference.}
\label{fig:semantic_masking}
\end{figure}

\subsection{Maximum-Intensity Tensor Obscuration}
Upon calculating the precise bounding parameters of the internal schematic relative to the pixel crop $T_{crop}$, the algorithm executes destructive obscuration. Literally translated, the algorithm completely wipes out all visual traces of the embedded figure from the table's image patch. Utilizing optimized OpenCV \cite{opencv} geometric convolution arrays (\texttt{cv2.rectangle}), the exact tensor zone corresponding to the graphical image is blasted with maximum-intensity pixel limits $(255, 255, 255)$, corresponding to true, opaque white in the RGB color space. 

For every sub-pixel coordinate $(x, y)$ residing within the calculated boundaries of the diagram, the original tensor intensity $T[x,y]$ is mathematically altered to true-white:

\begin{equation}
T_{crop}'(x,y) = 
\begin{cases} 
(255, 255, 255) & \text{if } x \in [x'_{min}, x'_{max}] \text{ and } y \in [y'_{min}, y'_{max}] \\
T_{crop}(x,y) & \text{otherwise}
\end{cases}
\end{equation}

\subsection{Deterministic Grid Resolution}
This mathematical intervention creates a pristine, functionally empty whitespace zone exactly where the dense engineering schematic previously resided. 

When the obscured tensor $T_{crop}'$ is evaluated dynamically by TATR, the model functions flawlessly. With the deceptive graphical vectors physically masked, the detection transformer correctly relies purely on the structural grid constraints of the remaining table to predict exact row, column, and cell coordinate lists. 

This highly localized engineering solution definitively eliminates topological table hallucination entirely independent of unreliable LLM prompting. Conclusively, it enables the integration of state-of-the-art TSR grid mapping for industrial documents, populating structurally verified topological trees directly into the unified Markdown abstraction.

\section{The Hierarchical Knowledge Graph (HKG) Construction}
\label{sec:hkg_logic}

Once layout regions have been isolated and their internal content extracted, the pipeline assembles them into a \textit{Hierarchical Knowledge Graph} (HKG). Unlike a flat Markdown file, the HKG preserves the multi-modal "connective tissue" of the document, ensuring that every extracted data point carries its full semantic context. The construction follows a three-pass algorithmic approach.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.2cm,
    node/.style={circle, draw, thick, align=center, minimum size=1.5cm, fill=white},
    edge/.style={->, thick, >=stealth}
]

% Main Graph Structure
\node[node, fill=gray!10] (doc) {Document\\Root};
\node[node, below=of doc, fill=blue!10] (page) {Page\\Node};
\node[node, below left=of page, fill=orange!10] (section) {Section\\Header};
\node[node, below right=of page, fill=green!10] (table) {Table\\Region};
\node[node, below=of section, fill=orange!5] (para) {Text\\Chunk};
\node[node, below=of table, fill=green!5] (row) {Data\\Record};

% Edges
\draw[edge] (doc) -- (page);
\draw[edge] (page) -- (section);
\draw[edge] (page) -- (table);
\draw[edge] (section) -- (para);
\draw[edge] (table) -- (row);

% Breadcrumb flows
\draw[edge, dashed, red, bend right] (section.west) to node[midway, left] {Context Injection} (para.west);
\draw[edge, dashed, red, bend left] (section) to node[midway, above] {Breadcrumb} (table);

\end{tikzpicture}
\caption{The schema of the Hierarchical Knowledge Graph (HKG). Dashed lines represent the "Breadcrumb" propagation logic where section context is injected into spatially or logically adjacent data chunks.}
\label{fig:hkg_structure}
\end{figure}

\subsection{Pass 1: Atomic Region Chunking}
The first pass maps the semantic bounding boxes derived from DocLayout-YOLO into atomic graph nodes. Each node stores the raw text, the region classification (e.g., \texttt{Table}, \texttt{Heading}), and its precise geometric provenance within the source PDF. This pass establishes the "spatial foundation" of the document.

\subsection{Pass 2: Spatial Linking and Caption Association}
In the second pass, the pipeline executes a distance-weighted linking algorithm to associate figures and tables with their respective captions. Industrial documents often place captions below tables or above figures; the graph builder computes the vertical distance between \texttt{Caption} nodes and their nearest \texttt{Table}/\texttt{Figure} neighbors. If the distance is within an adaptive threshold and no other semantic boundaries (like a section title) intervene, a \texttt{CAPTIONS\_FOR} edge is instantiated in the graph.

\subsection{Pass 3: Cross-Reference Resolution and Breadcrumbs}
The final pass implements the \textit{Breadcrumb System}. As the graph builder traverses the hierarchical tree, it identifies \texttt{Heading} and \texttt{Title} nodes. The text content of these headers is recursively propagated down to all "descendant" nodes (text paragraphs and tables) until the next header of equal or higher rank is encountered. 

This ensures that a data record extracted from a table on page 40 carries the semantic context of the section it belongs to (e.g., "Section 4.2: Technical Specifications"). To concretely illustrate this hierarchical structure, a resulting node generated dynamically by the HKG engine is serialized into strict JSON as follows:

\begin{verbatim}
{
  "node_id": "Table_Complex_1",
  "type": "Table",
  "content": { ... Markdown payload ... },
  "metadata": {
    "bbox": [120.5, 450.2, 600.0, 800.8],
    "page_num": 14
  },
  "edges": [
    {"target": "Caption_12", "relation": "CAPTIONS_FOR"}
  ],
  "breadcrumbs": ["Section 4: Technical Specs", "Operating Parameters"]
}
\end{verbatim}

\section{Graph-Native Markdown Ingestion}
\label{sec:markdown_assembly}

The convergence of the HKG and the extraction layers is finalized by the \textit{ConverterEngine}, which renders the graph into structured Markdown. Crucially, the engine does not just output text; it embeds hidden HTML provenance comments that map every Markdown block back to its unique Graph Node ID:

\begin{verbatim}
<!-- id=region_42 -->
### Supply Voltage
The module operates at 24V DC.
\end{verbatim}

This mapping allows downstream RAG applications to perform a "Graph-to-Text" lookup. When an LLM retrieves a specific text chunk, the RAG system can query the Knowledge Graph for the chunk's parent breadcrumbs, associated image captions, or neighboring tables, drastically reducing hallucination by providing the model with a complete semantic neighborhood rather than an isolated fragment of text. Furthermore, the integrity of these atomic Markdown blocks ensures that the \textit{ExtractorAgent} (described in Chapter \ref{ch:routing_vlm}) can reliably enforce its dynamic Pydantic schemas, successfully linking geometric PDF primitives to highly structured JSON output.
